A Transactional Memory System for Parallel Python

### 2014-08-02 talk/dls2014/paper/paper.tex
fix formatting and references

@@

\bibitem{cfbolz09} Carl Friedrich Bolz, Antonio Cuni, Maciej
  Fijalkowski, and Armin Rigo. 2009. Tracing the meta-level: PyPy's
  tracing JIT compiler.  \emph{In Proceedings of the 4th workshop on the
    Implementation, Compilation, Optimization of Object-Oriented Languages
    and Programming Systems} (ICOOOLPS '09).

\bibitem{kevin10} Kevin Millikin, Florian Schneider. 2010.  A New
  Crankshaft for V8.
  \url{http://blog.chromium.org/2010/12/new-crankshaft-for-v8.html}

@@

\bibitem{wayforward14} Remigius Meier, Armin Rigo. 2014. A Way Forward
  in Parallelising Dynamic Languages. To appear in ICOOOLPS'14.

@@

\bibitem{beazley10} Beazley, David. "Understanding the Python GIL."
  \emph{PyCON Python Conference}. Atlanta, Georgia. 2010.

\bibitem{harris10} Harris, Tim, James Larus, and Ravi
  Rajwar. "Transactional memory." \emph{Synthesis Lectures on Computer
  Architecture 5.1} (2010)

\bibitem{guerraoui08}
  Rachid Guerraoui and Michal Kapalka. 2008. On the correctness of
  transactional memory. In \emph{Proceedings of the 13th ACM SIGPLAN
    Symposium on Principles and practice of parallel programming} (PPoPP
  '08).

\bibitem{blundell06} Blundell, Colin, E. Christopher Lewis, and Milo
  Martin. "Unrestricted transactional memory: Supporting I/O and system
  calls within transactions." (2006).

\bibitem{spear08} Michael F. Spear and Michael Silverman and Luke
  Daless and Maged M. Michael and Michael L. Scott. 2008. "Implementing
  and exploiting inevitability in software transactional memory."  In
  \emph{Proc. 37th IEEE international conference on parallel
    processing} (ICPP '08), 59-66

\bibitem{fergus02} Fergus Henderson. 2002. Accurate garbage collection
  in an uncooperative environment. \emph{In Proceedings of the 3rd
    international symposium on Memory management} (ISMM '02).

\bibitem{stmupdate13} Armin Rigo, Remigius Meier. 2013. Update on
  STM. \url{morepypy.blogspot.ch/2013/10/update-on-stm.html}

\bibitem{rajwar05} Ravi Rajwar, Maurice Herlihy, and Konrad
  Lai. 2005. Virtualizing Transactional Memory. In \emph{Proceedings of
    the 32nd annual international symposium on Computer Architecture}
  (ISCA '05).

\bibitem{chung06} JaeWoong Chung, Chi Cao Minh, Austen McDonald,
  Travis Skare, Hassan Chafi, Brian D. Carlstrom, Christos Kozyrakis,
  and Kunle Olukotun. 2006. Tradeoffs in transactional memory
  virtualization. \emph{SIGOPS Oper. Syst. Rev.} 40, 5 (October 2006),
  371-381.

\bibitem{martin09} Martín Abadi, Tim Harris, and Mojtaba
  Mehrara. 2009. Transactional memory with strong atomicity using
  off-the-shelf memory protection hardware. \emph{SIGPLAN Not.} 44, 4
  (February 2009), 185-196.

\bibitem{pypybenchs} PyPy benchmarks repository. 2014. Revision
  a26f2fb58413. \url{bitbucket.org/pypy/benchmarks}

@@

\bibitem{odaira14}
  Rei Odaira, Jose G. Castanos, and Hisanobu Tomari. 2014. Eliminating
  global interpreter locks in Ruby through hardware transactional
  memory. In \emph{Proc. 19th ACM SIGPLAN Symposium on
    Principles and Practice of Parallel Programming} (PPoPP '14).

\bibitem{warmhoff13}
  Jons-Tobias Wamhoff, Christof Fetzer, Pascal Felber, Etienne Rivière,
  and Gilles Muller. 2013. FastLane: improving performance of software
  transactional memory for low thread counts. \emph{SIGPLAN Not.} 48, 8
  (February 2013), 113-122.

\bibitem{drago11} Aleksandar Dragojević, Pascal Felber, Vincent
  Gramoli, and Rachid Guerraoui. 2011. Why STM can be more than a
  research toy. \emph{Commun. ACM} 54, 4 (April 2011), 70-77.

\bibitem{cascaval08}
  Calin Cascaval, Colin Blundell, Maged Michael, Harold W. Cain, Peng
  Wu, Stefanie Chiras, and Siddhartha Chatterjee. 2008. Software
  transactional memory: why is it only a research
  toy?. \emph{Commun. ACM} 51, 11 (November 2008), 40-46.

\bibitem{nicholas06}
  Nicholas Riley and Craig Zilles. 2006. Hardware transactional memory
  support for lightweight dynamic language evolution. \emph{In
    Companion to the 21st ACM SIGPLAN symposium on Object-oriented
    programming systems, languages, and applications} (OOPSLA
  '06).

\bibitem{fuad10}
  Fuad Tabba. 2010. Adding concurrency in Python using a commercial
  processor's hardware transactional memory support. \emph{SIGARCH
  Comput. Archit. News 38}, 5 (April 2010), 12-19.

\bibitem{felber07}
  Pascal Felber and Torvald Riegel and Christof Fetzer and Martin
  Süßkraut and Ulrich Müller and Heiko Sturzrehm. 2007. Transactifying
  applications using an open compiler framework. \emph{TRANSACT}, August
  (2007): 4-6.

@@

\bibitem{spear09}
  Luke Dalessandro, Dave Dice, Michael Scott, Nir Shavit, and Michael
  Spear. 2010. Transactional mutex locks. In \emph{Proceedings of the
    16th international Euro-Par conference on Parallel processing: Part
    II} (Euro-Par'10), Pasqua D'Ambra, Mario Guarracino, and Domenico
  Talia (Eds.). Springer-Verlag, Berlin, Heidelberg, 2-13.

\bibitem{lamport79}
  L. Lamport. 1979. How to Make a Multiprocessor Computer That Correctly
  Executes Multiprocess Programs. \emph{IEEE Trans. Comput.} 28, 9
  (September 1979), 690-691.

\bibitem{victor11}
  Victor Pankratius and Ali-Reza Adl-Tabatabai. 2011. A study of
  transactional memory vs. locks in practice. In \emph{Proc.
     twenty-third annual ACM Symposium on Parallelism in Algorithms
    and Architectures} (SPAA '11).

\bibitem{christopher10}
  Christopher J. Rossbach, Owen S. Hofmann, and Emmett
  Witchel. 2010. Is transactional programming actually
  easier?. \emph{Proc. 15th ACM SIGPLAN Symposium on
    Principles and Practice of Parallel Programming} (PPoPP '10).

\bibitem{tim03}
  Tim Harris and Keir Fraser. 2003. Language support for lightweight
  transactions. \emph{In Proceedings of the 18th annual ACM SIGPLAN
    conference on Object-oriented programing, systems, languages, and
    applications} (OOPSLA '03), 388-402.

\bibitem{tim05} Tim Harris, Simon Marlow, Simon Peyton Jones, and
  Maurice Herlihy. 2008. Composable memory transactions.
  \emph{Commun. ACM} 51, 8 (August 2008), 91-100.

\bibitem{shan08}
  Shan Lu, Soyeon Park, Eunsoo Seo, and Yuanyuan Zhou. 2008. Learning
  from mistakes: a comprehensive study on real world concurrency bug
  characteristics. \emph{SIGARCH Comput. Archit. News} 36, 1 (March 2008),
  329-339.

\bibitem{bennet10}
  Bennet Yee, David Sehr, Gregory Dardyk, J. Bradley Chen, Robert
  Muth, Tavis Ormandy, Shiki Okasaka, Neha Narula, and Nicholas
  Fullagar. 2010. Native Client: a sandbox for portable, untrusted x86
  native code. \emph{Commun. ACM} 53, 1 (January 2010), 91-99.

% \bibitem{leis14}
%   Leis, Viktor, Alfons Kemper, and Thomas Neumann. "Exploiting
%   Hardware Transactional Memory in Main-Memory Databases."
%   \emph{Proc. of ICDE}. 2014.

>>>>

\vspace{1cm} %% formatting hack

@@

\bibitem{cfbolz09} Carl Friedrich Bolz, Antonio Cuni, Maciej
  Fijalkowski, and Armin Rigo. 2009. ``Tracing the meta-level: PyPy's
  tracing JIT compiler.''  \emph{Proc. 4th workshop on the
    Implementation, Compilation, Optimization of Object-Oriented Languages
    and Programming Systems} (ICOOOLPS '09), 18-25

\bibitem{kevin10} Kevin Millikin, Florian Schneider. 2010.  ``A New
  Crankshaft for V8.''
  \url{http://blog.chromium.org/2010/12/new-crankshaft-for-v8.html}

@@

\bibitem{wayforward14} Remigius Meier, Armin Rigo. 2014. ``A Way Forward
  in Parallelising Dynamic Languages.'' In \emph{Proc. 9th International
    Workshop on Implementation, Compilation, Optimization of Object-Oriented
    Languages, Programs and Systems PLE} (ICOOOLPS '14)

@@

\bibitem{beazley10} David Beazley. ``Understanding the Python GIL.''
  \emph{PyCON Python Conference}. Atlanta, Georgia. 2010.

\bibitem{harris10} Tim Harris, James Larus, and Ravi
  Rajwar. 2010. ``Transactional memory.'' \emph{Synthesis Lectures on Computer
  Architecture 5.1}

\bibitem{guerraoui08}
  Rachid Guerraoui and Michal Kapalka. 2008. ``On the correctness of
  transactional memory.'' In \emph{Proc. 13th ACM SIGPLAN Symposium on
    Principles and practice of parallel programming} (PPoPP '08),
  175-184

\bibitem{blundell06} Colin Blundell, E. Christopher Lewis, and Milo
  Martin. 2006. ``Unrestricted transactional memory: Supporting I/O and system
  calls within transactions.''

\bibitem{spear08} Michael F. Spear and Michael Silverman and Luke
  Daless and Maged M. Michael and Michael L. Scott. 2008. ``Implementing
  and exploiting inevitability in software transactional memory.''  In
  \emph{Proc. 37th IEEE international conference on parallel
    processing} (ICPP '08), 59-66

\bibitem{fergus02} Fergus Henderson. 2002. ``Accurate garbage collection
  in an uncooperative environment.'' In \emph{Proc. 3rd
    International Symposium on Memory management} (ISMM '02), 150-156

\bibitem{stmupdate13} Armin Rigo, Remigius Meier. 2013. ``Update on
  STM.'' \url{morepypy.blogspot.ch/2013/10/update-on-stm.html}

\bibitem{rajwar05} Ravi Rajwar, Maurice Herlihy, and Konrad
  Lai. 2005. ``Virtualizing Transactional Memory.'' In \emph{Proc.
    32nd annual International Symposium on Computer Architecture}
  (ISCA '05), 494-505

\bibitem{chung06} JaeWoong Chung, Chi Cao Minh, Austen McDonald,
  Travis Skare, Hassan Chafi, Brian D. Carlstrom, Christos Kozyrakis,
  and Kunle Olukotun. 2006. ``Tradeoffs in transactional memory
  virtualization.'' In \emph{Proc. 12th international conference on
    Architectural support for programming languages and operating
    systems} (ASPLOS XII), 371-381

\bibitem{martin09} Martín Abadi, Tim Harris, and Mojtaba
  Mehrara. 2009. ``Transactional memory with strong atomicity using
  off-the-shelf memory protection hardware.'' In \emph{Proc. 14th ACM
    SIGPLAN symposium on Principles and practice of parallel
    programming} (PPoPP '09), 185-196

\bibitem{pypybenchs} PyPy benchmarks repository. 2014. Revision
  fd2da4da8f33. \url{bitbucket.org/pypy/benchmarks}

@@

\bibitem{odaira14}
  Rei Odaira, Jose G. Castanos, and Hisanobu Tomari. 2014. ``Eliminating
  global interpreter locks in Ruby through hardware transactional
  memory.'' In \emph{Proc. 19th ACM SIGPLAN Symposium on
    Principles and Practice of Parallel Programming} (PPoPP '14), 131-142

\bibitem{warmhoff13}
  Jons-Tobias Wamhoff, Christof Fetzer, Pascal Felber, Etienne
  Rivière, and Gilles Muller. 2013. ``FastLane: improving performance
  of software transactional memory for low thread counts.'' In
  \emph{Proc. 18th ACM SIGPLAN symposium on Principles and practice of
    parallel programming} (PPoPP '13), 113-122

\bibitem{drago11} Aleksandar Dragojević, Pascal Felber, Vincent
  Gramoli, and Rachid Guerraoui. 2011. ``Why STM can be more than a
  research toy.'' \emph{Commun. ACM} 54, 4 (April 2011), 70-77.

\bibitem{cascaval08}
  Calin Cascaval, Colin Blundell, Maged Michael, Harold W. Cain, Peng
  Wu, Stefanie Chiras, and Siddhartha Chatterjee. 2008. ``Software
  transactional memory: why is it only a research
  toy?.'' \emph{Commun. ACM} 51, 11 (November 2008), 40-46.

\bibitem{nicholas06}
  Nicholas Riley and Craig Zilles. 2006. ``Hardware transactional memory
  support for lightweight dynamic language evolution.'' \emph{In
    Companion to the 21st ACM SIGPLAN symposium on Object-oriented
    programming systems, languages, and applications} (OOPSLA
  '06), 998-1008

\bibitem{fuad10}
  Fuad Tabba. 2010. ``Adding concurrency in Python using a commercial
  processor's hardware transactional memory support.'' \emph{SIGARCH
  Comput. Archit. News 38}, 5 (April 2010), 12-19.

\bibitem{felber07}
  Pascal Felber and Torvald Riegel and Christof Fetzer and Martin
  Süßkraut and Ulrich Müller and Heiko Sturzrehm. 2007. ``Transactifying
  applications using an open compiler framework.'' \emph{TRANSACT}, August
  (2007): 4-6.

@@

\bibitem{spear09}
  Luke Dalessandro, Dave Dice, Michael Scott, Nir Shavit, and Michael
  Spear. 2010. ``Transactional mutex locks.'' In \emph{Proc.
    16th international Euro-Par conference on Parallel processing: Part
    II} (Euro-Par'10), Pasqua D'Ambra, Mario Guarracino, and Domenico
  Talia (Eds.). Springer-Verlag, Berlin, Heidelberg, 2-13.

\bibitem{lamport79}
  L. Lamport. 1979. ``How to Make a Multiprocessor Computer That Correctly
  Executes Multiprocess Programs.'' \emph{IEEE Trans. Comput.} 28, 9
  (September 1979), 690-691.

\bibitem{victor11}
  Victor Pankratius and Ali-Reza Adl-Tabatabai. 2011. ``A study of
  transactional memory vs. locks in practice.'' In \emph{Proc.
     23rd annual ACM Symposium on Parallelism in Algorithms
    and Architectures} (SPAA '11), 43-52

\bibitem{christopher10}
  Christopher J. Rossbach, Owen S. Hofmann, and Emmett
  Witchel. 2010. ``Is transactional programming actually
  easier?.'' \emph{Proc. 15th ACM SIGPLAN Symposium on
    Principles and Practice of Parallel Programming} (PPoPP '10), 47-56

\bibitem{tim03}
  Tim Harris and Keir Fraser. 2003. ``Language support for lightweight
  transactions.'' In \emph{Proc. 18th annual ACM SIGPLAN
    conference on Object-oriented programing, systems, languages, and
    applications} (OOPSLA '03), 388-402.

\bibitem{tim05} Tim Harris, Simon Marlow, Simon Peyton Jones, and
  Maurice Herlihy. 2008. ``Composable memory transactions.''
  \emph{Commun. ACM} 51, 8 (August 2008), 91-100.

\bibitem{shan08}
  Shan Lu, Soyeon Park, Eunsoo Seo, and Yuanyuan Zhou. 2008. ``Learning
  from mistakes: a comprehensive study on real world concurrency bug
  characteristics.'' \emph{SIGARCH Comput. Archit. News} 36, 1 (March 2008),
  329-339.

\bibitem{bennet10}
  Bennet Yee, David Sehr, Gregory Dardyk, J. Bradley Chen, Robert
  Muth, Tavis Ormandy, Shiki Okasaka, Neha Narula, and Nicholas
  Fullagar. 2010. ``Native Client: a sandbox for portable, untrusted x86
  native code.'' \emph{Commun. ACM} 53, 1 (January 2010), 91-99.

% \bibitem{leis14}
%   Leis, Viktor, Alfons Kemper, and Thomas Neumann. ``Exploiting
%   Hardware Transactional Memory in Main-Memory Databases.''
%   \emph{Proc. of ICDE}. 2014.
